Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-05 Manuscript/manuscript.Rmd
edits

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, this is a multi-faceted probelem. Many methods ignore key characteristics [?? which ones ??; !!!editted] of 3D genomic data including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between minority (regions with TAD boundaries) and majority classes (regions without TAD boundaries), the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??; !!!].

We can account for imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. Evaluation level solutions include choosing suitable performance metrics for evaluating models. <!--These solutions are necessary given that the test set, used to validate re-sampled training sets, is itself imbalanced.--> It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has researched previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

Furthermore, determing the optimal representation of the relationship between features in the model and genomic regions, known as feature engineering, plays a significant role in TAD prediction. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A lower resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances. To date, no study has compared the results of their predictive models built across multiple data resolutions.

@@

In this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation [?? Not sure what spatial representation means in this context ??] linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. [?? It feels like some stronger statements need to be here. I would make a statement about predicting TAD regions using genomic annotations instead of using terms like "linking". It feels like the focus should be on this and not model selection which occurs in any analysis ??]

For each resolution, we examined the inclusion of three different predictor types [?? Maybe list all 3 (.e.g "For each resolution, we examined the inclusion of three different predictor types: counts, percentages and distances.") ??], including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. This process was performed on chromosome-specific regions on the linear genome. Noninformative features were removed using a feature reduction algorithm. Lastly, random forest classifiers were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

>>>>

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, this is a multi-faceted probelem. Many methods ignore key characteristics [?? which ones ??; !!!editted] of 3D genomic data. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->

Firstly, sparsity of TAD boundary regions throughout the linear genome contribute to heavily imbalanced classes. This imbalance results in small imbalance ratios. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??; !!!].

We can account for this imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. Evaluation level solutions include choosing suitable performance metrics for evaluating models. <!--These solutions are necessary given that the test set, used to validate re-sampled training sets, is itself imbalanced.--> It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has been studied previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

Secondly, it is important to determine the optimal representation of the relationship between features in the model and genomic regions. This is known as feature engineering. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e.g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

Lastly, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A lower resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances. To date, no study has compared the results of their predictive models built across multiple data resolutions.

@@

In this study, we developed a novel ensemble framework aimed at systematically comparing the performances of random forest classification models in order to determine the set of data level characteristics necessary for optimal prediction of TAD boundaries. We evaluated performances of Hi-C data for both the GM12878 and K562 cell line obtained at four different Hi-C resolutions including 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each cell line and resolution, we compared three feature engineering techniques: counts, percents, and distances. Additionally, we compared three re-sampling techniques: random under-sampling, random-over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE).

[???Include concluding paragraph]
